Stack Safety as a Security Property

### 2020-07-08 main.tex
pass through 4; releasing lock on 4

\section{Threat Model}

@@

\bcp{More explanation needed, plus examples.}
A policy is a general model of an enforcement mechanism, consisting of a
set of policy states \(\pol \in \POLS\), a policy step function
\((\mach, \pol) \stepstopol \pol' : \MACHS \times \POLS
\rightharpoonup \POLS\), and a initializer \(\pinit : \MACHS \rightharpoonup
\POLS\). The initializer establishes the initial policy state for a given
machine state, but is partial to reflect that some policies are only defined
on states that are well-formed in some way.

\leo{TODO: Forward pointer to policies right?}\bcp{Yes.  But also we can say
in more detail that the intention is to capture any kind of dynamic
enforcement mechanism and give as many names and citations as we can.}

@@

In this section we describe our machine model. Our definition of stack
safety is generic and this section discusses an abstract interface to
our machine model and some kind of static or dynamic, software- or
hardware-enforced security monitor. We will refer back to the running
example of the previous section for concreteness, and then,
in~\cref{sec:enforcement}, extend this model with a dynamic policy
enforcement mechanism based on
\citet{DBLP:conf/sp/RoesslerD18}.

@@

The basic building blocks of the machine are {\em values} and {\em
  addresses}. Both are drawn from some set of {\em words}
$\WORDS$, ranged over by $\word$.
%
Our machine states are composed of {\em components} $k$, addressed either by
words (memory addresses) or by by register names ($r$) drawn from some set
$\REGS$.
%
    \[\component \in \COMPONENTS = \WORDS + \REGS \]
%
The register names are assumed to include two special purpose registers: the
program counter {\PCname} and the stack pointer \SP.

@@

The step function of the machine takes a machine state and produces a
pair of a machine state and an {\em observation}, which may be either
silent (written $\tau$) or a word of data that the machine
communicates to the outside world (e.g., by writing to a DMA
address):
%
\[\obs \in \OBSS = \{\tau\} \uplus \WORDS\]
\[\mach \stepstoobs{\obs} \mach' : \MACHS \rightarrow \MACHS \times \OBSS \]
%

The choice of observations is an interesting one. A first approach
would be to consider any change to the registers or memory as observable.
However, as some parts of the stack will be considered ``private''
this choice won't work. A relaxation of this to consider all changes
to ``non-private'' data as observable could work in principle, but
does not allow for an interesting weaker-but-more-efficient enforcement
mechanism. We will come back to this point in \Cref{sec:lazy}. On the
other hand, we could use a particular location as an abstract output
of our system (e.g., through DMA). This is the approach we chose for
our running example to keep things simple. For that, $\WORDS$ is the
set of 64-bit integers, $\REGS$ is the set $\{r_0, \cdots, r_{31}\}
\cup \{\PCname\}$, and $\SP = r_2$. Observations are the writes to the
special $\rout$ location: if during an execution step a word $w$ is
written to $\rout$ then $\obs = w$, otherwise $\obs = \tau$.

@@

In our model, the machine step function is total -- a program does not halt
on its own, but may instead enter a silent terminal loop. Runs are only cut
short by the policy step function. This means that software-only policies that
can catch errors must be modeled as reaching an error state, not outright
halting.

@@

In a similar fashion, we define \(\mpstatename \hookrightarrow_P \MPT\)
to relate a machine-policy state pair to the MP-trace
induced by \(\stepstopol{}\). Unlike the $\MOTRACE$ above, this trace can
be finite as the policy can cause a fault, in which case the final state is
a \(\tau\):
\rb{This $\hookrightarrow_P$ is often enough written just $\hookrightarrow$.
Overload everywhere?}

\begin{minipage}{.4\textwidth}
\judgment{\(\neg \exists \pol_1 . \pol_0 \rightharpoonup \pol_1\)}
         {\(\mpstate{\mach_0}{\pol_0} \hookrightarrow_p \mpostate{\mach_0}{\pol_0}{\tau}\)}
\end{minipage}
\begin{minipage}{.55\textwidth}
  \judgmenttwo{\(\mpstate{\mach_0}{\pol_o} \stepstoobs{\obs}
                 \mpstate{\mach_1}{\pol_1}\)}
              {\(\mpstate{\mach_1}{\pol_1} \hookrightarrow_P \MPT\)}
              {\(\mpstate{\mach_0}{\pol_0} \hookrightarrow_P
                 \mpostate{\mach_0}{\pol_0}{\obs} \MPT\)}
\end{minipage}

@@

\paragraph*{Until}
The ``Until'' operation \(\PUT(\Trace,f)\)
%\bcp{Wouldn't
%``While'' be more mnemonic?} \sna{Appealing, but it's more of a
%``while-not''? Trying out ``until''. In any case... moving away
%from doing joins means this needs to get some work to eliminate
%the trailing observation. Here's an attempt.}
takes a trace
\(\Trace : \mathit{trace} ~ \TRACEELEMS \times \OBSS\) and a predicate
on elements \(f \subset \TRACEELEMS\) and gives the prefix of
\(\Trace\) ending with the first element on which \(f\) holds.
Subtly, because the observation associated with a trace represents its
next step, when we take a prefix from a trace containing observations
we must replace the final observation with a \(\tau\).

  \begin{center}
  \begin{minipage}{.3\textwidth}
    \judgment{\(f\ \traceelem\)}
             {\(\PUT(f,(\traceelem,\obs)\Trace) = (\traceelem,\tau)\)}
  \end{minipage}
%
  \begin{minipage}{.3\textwidth}
  \judgment{}
           {\(\PUT(f,(\traceelem,\obs)) = (\traceelem,\tau)\)}
%
  \end{minipage}
  \begin{minipage}{.3\textwidth}
  \judgmenttwo{\(\neg f\ \traceelem\)}{\(\PUT(f,\Trace) = \Trace'\)}
              {\(\PUT(f,(\traceelem,\obs)\Trace) = (\traceelem,\obs)\Trace'\)}
  \end{minipage}
  \end{center}
%
Note that if the resulting trace is a strict prefix of the parameter, then the
predicate must hold on its final element. If the predicate never holds,
\(\PUT\) is the identity function.

@@

\judgmenttwo{\(\mach \hookrightarrow \machT'\)}
            {\(\PUT(f,\machT')
              = \machT\)}
            {\(\mach \hookrightarrow \machT | f\)}
\noindent
And similarly for traces with policies:

\judgmenttwo{\(\mpstatename \hookrightarrow \MPT'\)}
            {\(\PUT(\lambda \mpstate{\mach}{\pol}.f\ \mach,\MPT') = \MPT\)}
         {\(\mpstatename \hookrightarrow \MPT | f\)}

@@

We say that a trace of observations $\obsT_1$ is a prefix of $\obsT_2$
as far as an external observer is concerned, if the sequence of
non-silent observations of $\obsT_1$ is a prefix of those of
$\obsT_2$; that is we operate up to deletion of \(\tau\) observations,
coinductively:

@@

\leo{I still don't know how to do this transition. Maybe contrast
with the ``perhaps surprisingly'' later?} Note that
similarity of observation traces could alternatively be defined coinductively,
just like the prefix relation, but by dropping the assymetric rules that allow
for a finite trace ($\tau \lesssim \obsT$ or $w \lesssim w\obsT$)

In addition, perhaps surprisingly, an infinite silent trace is a
prefix of (and equivalent to) any other trace. While this might seem
surprising at first, it makes sense in a timing-insensitive context:
an external observer looking at two machine runs cannot (computably)
distinguish between a machine that steps forever and a machine that
steps for a long time before producing some output.

>>>>

\section{Threat Model}
\label{sec:threat}

@@

\apt{Informal definition of policies as auxiliary state here.}

@@

In this section we describe our machine model. Our definition of stack
safety is generic and this section discusses an abstract interface to
our machine model and some kind of static or dynamic, software- or
hardware-enforced security monitor. We will refer back to the running
example of the previous section for concreteness, and then,
in~\cref{sec:enforcement}, extend this model with a dynamic policy
enforcement mechanism based on
\citet{DBLP:conf/sp/RoesslerD18}.\apt{needs a relook}

@@

The basic building blocks of the machine are {\em values} and {\em
  addresses}. Both are drawn from some set of {\em words}
$\WORDS$, ranged over by $\word$.
%
Our machine states are composed of {\em components} $k$, addressed either by
words (memory addresses) or by register names ($r$) drawn from some set
$\REGS$.
%
    \[\component \in \COMPONENTS = \WORDS + \REGS \]
%
The register names are assumed to include two special purpose registers: the
program counter {\PCname} and the stack pointer \SP.

@@

The step function of the machine takes a machine state and produces a
pair of a machine state and an {\em observation}, which may be either
silent (written $\tau$) or a word of data that the machine
communicates to the outside world:
%
\[\obs \in \OBSS = \{\tau\} \uplus \WORDS\]
\[\mach \stepstoobs{\obs} \mach' : \MACHS \rightarrow \MACHS \times \OBSS \]
%

Notice that we choose to model observations explicitly, rather than simply treating
a subset of registers or memory as being observable.  This choice gives
us the flexibility to treat programs as observationally equivalent even when
their internal states may differ.
%The choice of modelling observations is an interesting one. A first approach
%would be to consider any change to the registers or memory as observable.
%However, as some parts of the stack will be considered ``private''
%this choice won't work. A relaxation of this to consider all changes
%to ``non-private'' data as observable could work in principle, but
%does not allow for an interesting weaker-but-more-efficient enforcement
%mechanism. We will come back to this point in \Cref{sec:lazy}.

In the running example of \Cref{sec:running-example}, 
$\WORDS$ is the
set of 64-bit integers and $\REGS$ is the set $\{r_0, \cdots, r_{31}\}
\cup \{\PCname,\SP\}$. Observations are the writes to the
  special $\rout$ location, which can be thought of as a memory-mapped
  output port: if during an execution step a word $w$ is
written to $\rout$ then $\obs = w$, otherwise $\obs = \tau$.

\subsection{Policies}

A policy is a general model of an enforcement mechanism, consisting of a
set of policy states \(\pol \in \POLS\), a policy step function
\((\mach, \pol) \polstep \pol' : \MACHS \times \POLS
\rightharpoonup \POLS\), and an initializer \(\pinit : \MACHS \rightharpoonup
\POLS\). The initializer establishes the initial policy state for a given
machine state, but is partial to reflect that some policies are only defined
on states that are well-formed in some way.\apt{Is this our current thinking?}

\apt{TODO: relate back to \Cref{sec:threat} and forward to \Cref{sec:enforcement}.}
%\leo{TODO: Forward pointer to policies right?}\bcp{Yes.  But also we can say
%in more detail that the intention is to capture any kind of dynamic
%enforcement mechanism and give as many names and citations as we can.}

@@

In our model, the machine step function is total -- a program does not halt
on its own, but may instead enter a silent terminal loop. Runs are only cut
short by the policy step function. This means that software-only policies that
can catch errors must be modeled as reaching an error state\apt{??}, not 
halting outright.\apt{Is this our current thinking?}

@@

In a similar fashion, we define \(\mpstatename \hookrightarrow \MPT\)
to relate a machine-policy state pair to the MP-trace
induced by the step function on $\MPS$. Unlike the $\MOTRACE$ above, this trace can
be finite as the policy can cause a fault, in which case the final observation is
a \(\tau\):
\rb{This $\hookrightarrow_P$ is often enough written just $\hookrightarrow$.
  Overload everywhere?}\apt{Done, and for $\rightharpoonup$ too.}

\begin{minipage}{.4\textwidth}
\judgment{\(\neg \exists \pol_1 . \mpstate{\mach_0}{\pol_0} \polstep \pol_1\)}
         {\(\mpstate{\mach_0}{\pol_0} \hookrightarrow \mpostate{\mach_0}{\pol_0}{\tau}\)}
\end{minipage}
\begin{minipage}{.55\textwidth}
  \judgmenttwo{\(\mpstate{\mach_0}{\pol_o} \stepstopol{\obs}
                 \mpstate{\mach_1}{\pol_1}\)}
              {\(\mpstate{\mach_1}{\pol_1} \hookrightarrow \MPT\)}
              {\(\mpstate{\mach_0}{\pol_0} \hookrightarrow
                 \mpostate{\mach_0}{\pol_0}{\obs} \MPT\)}
\end{minipage}

@@

\paragraph*{Until}
The operation \(\PUT~(\Trace,f)\)
%\bcp{Wouldn't
%``While'' be more mnemonic?} \sna{Appealing, but it's more of a
%``while-not''? Trying out ``until''. In any case... moving away
%from doing joins means this needs to get some work to eliminate
%the trailing observation. Here's an attempt.}
takes a trace
\(\Trace : \mathit{trace} ~ \TRACEELEMS \times \OBSS\) and a predicate
on elements \(f \subset \TRACEELEMS\) and gives the prefix of
\(\Trace\) ending with the first element on which \(f\) holds.
Subtly, because the observation associated with a trace represents its
next step, when we take a prefix from a trace containing observations
we must replace the final observation with a \(\tau\).

  \begin{center}
  \begin{minipage}{.3\textwidth}
    \judgment{\(f\ \traceelem\)}
             {\(\PUT~(f,(\traceelem,\obs)\Trace) = (\traceelem,\tau)\)}
  \end{minipage}
%
  \begin{minipage}{.3\textwidth}
  \judgment{}
           {\(\PUT~(f,(\traceelem,\obs)) = (\traceelem,\tau)\)}
%
  \end{minipage}
  \begin{minipage}{.3\textwidth}
  \judgmenttwo{\(\neg f\ \traceelem\)}{\(\PUT~(f,\Trace) = \Trace'\)}
              {\(\PUT~(f,(\traceelem,\obs)\Trace) = (\traceelem,\obs)\Trace'\)}
  \end{minipage}
  \end{center}
%
Note that if the resulting trace is a strict prefix of the parameter, then the
predicate must hold on its final element. If the predicate never holds,
\(\PUT\) is the identity function.

@@

\judgmenttwo{\(\mach \hookrightarrow \machT'\)}
            {\(\PUT~(f,\machT')
              = \machT\)}
            {\(\mach \hookrightarrow \machT | f\)}
\noindent
And similarly for traces with policies:

\judgmenttwo{\(\mpstatename \hookrightarrow \MPT'\)}
            {\(\PUT~(\lambda \mpstate{\mach}{\pol}.f\ \mach,\MPT') = \MPT\)}
         {\(\mpstatename \hookrightarrow \MPT | f\)}

@@

We say that a trace of observations $\obsT_1$ is a prefix of $\obsT_2$
as far as an external observer is concerned, if the sequence of
non-silent observations of $\obsT_1$ is a prefix of those of
$\obsT_2$; that is, we operate up to deletion of \(\tau\) observations,
coinductively:

@@

\leo{I still don't know how to do this transition. Maybe contrast
with the ``perhaps surprisingly'' later?} Note that
similarity of observation traces could alternatively be defined coinductively,
just like the prefix relation, but by dropping the asymmetric rules that allow
for a finite trace ($\tau \lesssim \obsT$ or $w \lesssim w\obsT$).\apt{Is this immportant?}

Note that an infinite silent trace is a
prefix of (and equivalent to) any other trace. While this might seem
surprising at first, it makes sense in a timing-insensitive context:
an external observer looking at two machine runs cannot (computably)
distinguish between a machine that steps forever and a machine that
steps for a long time before producing some output.
